Topic modeling on EU and USA research projects

### 2017-06-28 thesis_report/draft_thesis.tex
Final report

\title{Topic modeling on EU and the USA research projects}
\author{Maria Iliadi}

@@

\begin{titlepage}
	\centering
	\includegraphics[width=0.8\textwidth]{figs/logo.jpg}\par\vspace{4cm}
	{\scshape\Large MSc thesis\par}
	\vspace{1.5cm}
	{\Large\bfseries Topic modeling on EU and the USA research projects\par}
	\vspace{2cm}
	{\Large Maria Iliadi\par}
	\vfill
	supervised by\par
	Assoc.Prof.~Panos Louridas

@@

\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
First of all, I would like to express my gratitude to my supervisor, Panos
Louridas. Panos provided me with invaluable advice and help throughout my MSc
thesis with his profound knowledge and a seemingly never-ending patience whilst
allowing me to work in my own way. Panos’ positive outlook and methodic approach
to solving problems has been a huge inspiration to me. He also helped me
comprehend the right research practice making these months of study a real
pleasure. Thank you.

Many thanks to my colleagues from the IST Lab. It was an honour to have a chance
to work together with such great people and brilliant researchers that helped me
broaden my research interests. I would also like to thank all of my colleagues
from the MSc in Computer Science and especially Artem Tsikiridis. I really
appreciated helpful discussions and much needed coffee breaks. Finally, I would
like to thank my family and my friends for their support and love all these
years.
\end{abstract}

@@

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
funds approximately 24\% of the federally supported research
conducted by the USA's colleges and universities\footnote{\url{https://www.nsf.gov/about/}}.

@@

In section \ref{theoretic} introduces some basic concepts
in topic modeling and we describe the model that we used for topic
extraction, Latent Dirichlet Allocation, which is a statistical
generative model for documents. In section \ref{relatedwork}, 
we briefly talk about the papers that mostly influenced our study. 
In section \ref{methodology} we provide information
about the dataset structure and the methodology that we followed in
our analysis. In section \ref{analysis} we interpret the results of 
our analysis, and we discuss the evolution of the topics over time and the
differences between the two regions' topics based on their similarity.
Finally, section \ref{conclusion} summarizes our work and section 
\ref{future} suggests some future work.

\section{Theoretic background}
\label{theoretic}
In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive texts body, machine learning 
researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents\cite{Blei11introductionto}.

@@

When introducing the Latent Dirichlet Allocation in section \ref{LDAexplained}, 
we will be using the terms prior and posterior probabilities
and likelihood function:

@@

where $\mathbb{E}$ is the expectation of the random variable $X$
\footnote{Suppose random variable $X$ can take value $x_{1}$ with 
probability $p_{1}$, value $x_{2}$ with probability $p_{2}$, 
and so on, up to value $x_{k}$ with probability $p_{k}$. 
Then the expectation of this random variable $X$ is defined as
$\mathbb{E}[X] = x_{1}p_{1} + x_{2}p_{2} + ... + x_{k}p_{k}$.}.

@@

The probability distribution that the model uses is the multinomial, 
because it assumes that this is the distribution of topics in 
documents and words in topics (read Latent Dirichlet Allocation in section \ref{LDAexplained}). 
In probability theory, the multinomial distribution is a generalization
of the binomial distribution. For example, it models the probability
of counts for rolling a $k$-sided dice $n$ times. For $n$ independent trials
each of which leads to a success for exactly one of $k$ categories, with
each category having a given fixed success probability, the
multinomial distribution gives the probability of any particular
combination of numbers of successes for the various
categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words\cite{Huang_maximumlikelihood}.

@@

The Dirichlet distribution is a way to model random probability mass functions
(PMFs)\footnote{If you roll 1000 dice, the theoretical odds of any particular
number showing up (i.e. a 1, 2, 3, 4, 5, or 6) are 1/6. However, you won't get
that exact distribution in a real experiment due to manufacturing defects. If
you have ten dice, each die will have its own probability mass function (PMF).}
for finite sets. It is also sometimes used as a prior in Bayesian statistics.
The Dirichlet is the multivariate generalization of the beta distribution. It
is an extension of the beta distribution for modeling probabilities for two or
more events; when the result of the event has only 2 values, the Dirichlet
distribution is equal to the beta distribution.

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990\cite{Deerwester90indexingby}. The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent
Semantic Indexing, or PLSI model\cite{Hofmann:1999:PLS:312624.312649}. 
PLSI is a topic model where each word in a document is generated 
from a single topic which results in that each document in a corpus 
can be represented with a topic distribution. Later, in 2003 Blei et al. 
presented the Latent Dirichlet Allocation, or LDA\cite{blei2003latent}. 
As opposed to PLSI, LDA is a statistically generative model for documents 
where each word in a document can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them\cite{Deerwester90indexingby}. In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF\footnote{TF-IDF stands 
for term frequency - inverse document frequency, and the TF-IDF weight is 
a weight often used in information retrieval and text mining. This weight 
is a statistical measure used to evaluate how important a word is to 
a document in a collection or corpus. The importance increases
proportionally to the number of times a word appears in the document (TF) but
is offset by the frequency of the word in the corpus (IDF).}. Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts\cite{Salton:1988:TAA:54259.54260}.

@@

\begin{enumerate}
\item Pick a document \textbf{w} with probability $p($\textbf{w}$)$
\item For each of the $N$ words in \textbf{w}:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|$\textbf{w}$)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/PLSI_plate_notation.png}
\caption{Plate notation representing the PLSI model. \textbf{w} is the document 
index, $z$ is the topic drawn for the word $w$ from the topic distribution of 
the document}
\end{figure}
\end{center}
The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|$\textbf{w}$)$. Even though PLSI is generative for 
the analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data\cite{blei2003latent}. Another problem is that the number 
of parameters in the model grows linearly with the number of documents.

\subsubsection{Latent Dirichlet Allocation (LDA)}
\label{LDAexplained}
LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics\cite{Blei11introductionto}.

A limitation with PLSI is that there is no generative model for these
proportions, making it difficult to handle unseen documents. The reduced
document description result in PLSI is a vector where each element describes the
mixing proportion for a topic. The Latent Dirichlet Allocation model tries to
solve this limitation by setting a Dirichlet prior on the topic
distribution\cite{blei2003latent}. In practice, this results in more reasonable
mixtures of topics in a document.

@@

The basic idea in LDA is that we define a topic to be a multinomial
distribution over a fixed vocabulary. Technically, the model assumes 
that the topics are generated first, before the
documents\cite{Blei11introductionto}. Now for each document in the
collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit
multiple topics. Each document exhibits the topics with different
proportion (step 1); each word in each document is drawn from one of
the topics (step 2b), where the selected topic is chosen from the
per-document distribution over topics (step 2a). All the documents in
the corpus share the same set of topics, but each document exhibits
those topics with different proportion - that's the distinguishing
characteristic of LDA. A more detailed explanation of the model
follows in section \ref{LDAmodel}.

@@

For each document $w$, in corpus $D$, the LDA generative process is as follows, 
with corresponding plate notation in Figure \ref{figLDA}:

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section \ref{oLDAvb}.

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously
\cite{Myung:2003}.

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use 
MCMC with Gibbs sampling\cite{Griffiths02gibbssampling} and expectation 
propagation\cite{Minka:2002:EGA:2073876.2073918}. The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed\cite{Fox2011ATO}.

@@

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets
\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

\paragraph{Online variational LDA}
\label{oLDAvb}
An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The \emph{online LDA} is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

We can see that unlike variational LDA, in online LDA we only need to
hold a small chunk of the data at a time, and once we're done
analyzing it, we never need it again. As with batch, once we've
estimated $\lambda$, we can find the most probable words for each
topic by looking at the word probabilities in each row of
$\lambda$\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

@@

D. Blei's research focuses on probabilistic topic models, Bayesian methods and
approximate posterior inference. In his work \emph{Introduction to
Probabilistic Topic Models}\cite{Blei11introductionto} 
he describes the basic ideas behind LDA and
applies the model to articles from the journal \emph{Science}. 
The model assumes that the topics are generated
first, before the documents and that the documents exhibit multiple
topics, but each one of them exhibit those topics in different
proportion. In our implementation we also use the LDA model and we
follow a similar approach, since we don't use any previous information
about the subjects that run through our dataset and the documents are
not labeled with keywords or topics. The topic distributions arise by
computing the hidden structure that resembles the thematic structure
of the collection.

The paper that mostly influenced our work is \emph{Finding Scientific Topics}, by
T. Griffiths and M. Steyvers\cite{griffiths_steyvers04}. They investigated
abstract data from articles published in the Proceedings of the National Academy
of Sciences (PNAS) from 1991 to 2001 and compared research topics obtained from
topic modeling with existing categories. In our implementation, we also use 
the abstracts of the projects as documents for the model.

@@

Inspired by Griffiths and Steyvers, L. Sun and Y. Yin wrote a paper,
\emph{Discovering themes and trends in transportation research using topic
modeling}\cite{Sun201749}, about topic modeling in transportation
research. Considering approximately 17000 abstracts of articles
published on 22 leading journals in transportation research as
documents, they infer 50 topics by using LDA. Compared to \emph{Finding
scientific topics}, Sun and Yin use more than one journal (22 popular
journals) and they investigate the journal-topic distribution and
journal similarity. In our analysis, we identify the EU and the USA as
2 regions (like 2 different journals), with different proportions of
documents-projects, and we visualize the topic similarity of the two
regions per framework programme.

@@

\section{Methodology}
\label{methodology}
In this section, we explain the procedure that we followed in order to
apply the LDA model in our dataset. It consists of information about 
EU projects funded for research and innovation under the FP4-H2020 
and the awards/projects that got funded by NSF from 1994-2017 in USA. We
consider the projects' description/abstract as the documents in LDA.

@@

Below, we provide the general topics of the EU projects as listed in the EU 
Open Data Portal \footnote{https://data.europa.eu/euodp/en/data}:
\begin{itemize}
\item Agriculture, Fisheries and Food
	\begin{description}
	\item Agriculture, Food safety, Maritime affairs and fisheries
	\end{description}
\item Business
	\begin{description}
	\item Competition, Enterprise, Single market, Trade
	\end{description}
\item Culture and Education
	\begin{description}
	\item Audiovisual and media, Culture, Education training and youth, Multilingualism
	\end{description}
\item Customs and Tax
	\begin{description}
	\item Customs, Taxation
	\end{description}
\item Development and Humanitarian aid
	\begin{description}
	\item Development and cooperation, Humanitarian aid and Civil protection, Human rights
	\end{description}
\item Economy and Finance
	\begin{description}
	\item Budget, Economic and monetary affairs, Fraud prevention
	\end{description}
\item Employment and Social affairs
\item Enlargement and Foreign affairs
\item Environment and Energy
	\begin{description}
	\item Climate action, Energy, Environment
	\end{description}
\item EU institutions
	\begin{description}
	\item Institutional affairs
	\end{description}
\item Health
	\begin{description}
	\item Health, Sport
	\end{description}
\item Justice and Citizens' rights
	\begin{description}
	\item EU citizenship, Consumers, Justice and Home affairs
	\end{description}
\item Regions and Local development
	\begin{description}
	\item Regional policy
	\end{description}
\item Science and Technology
	\begin{description}
	\item Space, Digital Economy and Society, Research and Innovation
	\end{description}
\item Transport and Travel
\end{itemize}

The data that are provided by CORDIS (via the EU Open Data Portal) are
open, free and can be used in research, applications, commercial or
non-commercial purposes. The dataset that we chose for the following
analysis contains information about all the projects funded by the
European Union from 1994 to 2017. These projects got approved under
the framework programme (FP4-H2020) for research and technological
development. Grant information is provided for each project, 
including reference, acronym, dates, funding, programmes, 
participant countries, subjects and objectives.

@@

\begin{itemize}
\item FP4: fourth framework programme (1994-1998)
\item FP5: fifth framework programme (1998-2002)
\item FP6: sixth framework programme (2002-2006)
\item FP7: seventh framework programme (2007-2013)
\item H2020: Horizon 2020 framework programme (2014-2020)
\end{itemize}

@@

\emph{Management software \& systems} is a subject that appears only in FP4 and
FP5, with the 12\% of the projects. emph{Information systems \& technologies'}
related projects cover the 8,5\% of the projects on average from FP5-H2020.
Since technology has an impact in every aspect nowadays, it is normal to have a
stable popularity over the years.

>>>>

\title{Topic modeling on EU and USA research projects}
\author{Maria Iliadi}

@@

\begin{titlepage}
	\centering
	\includegraphics[width=0.8\textwidth]{figs/logo.jpg}\par\vspace{4cm}
	{\scshape\Large MSc thesis\par}
	\vspace{1.5cm}
	{\Large\bfseries Topic modeling on EU and the USA research projects\par}
	\vspace{2cm}
	{\Large Maria Iliadi\par}
	\vfill
	supervised by\par
	Assoc. Prof.~Panos Louridas

@@

\renewcommand{\abstractname}{Acknowledgments}
\begin{abstract}
First of all, I would like to express my gratitude to my supervisor, Panos
Louridas. Panos provided me with invaluable advice and help throughout my MSc
thesis with his profound knowledge and a seemingly never-ending patience whilst
allowing me to work in my own way. Panos’s positive outlook and methodic approach
to solving problems has been a huge inspiration to me. He also helped me
comprehend the right research practice making these months of study a real
pleasure. Thank you.

Many thanks to my colleagues from the IST Lab. It was an honor to have a chance
to work together with such great people and brilliant researchers that helped me
broaden my research interests. I would also like to thank all of my colleagues
from the MSc in Computer Science and especially Artem Tsikiridis. I really
appreciated helpful discussions and much needed coffee breaks. Finally, I would
like to thank my family and my friends for their support and love all these
years.
\end{abstract}

@@

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
funds approximately 24\% of the federally supported research
conducted by the USA's colleges and universities~\footnote{\url{https://www.nsf.gov/about/}}.

@@

In section~\ref{theoretic} we introduce some basic concepts
in topic modeling and we describe the model that we used for topic
extraction, Latent Dirichlet Allocation, which is a statistical
generative model for documents. In section~\ref{relatedwork}, 
we briefly talk about the papers that mostly influenced our study. 
In section~\ref{methodology} we provide information
about the dataset structure and the methodology that we followed in
our analysis. In section~\ref{analysis} we interpret the results of 
our analysis, and we discuss the evolution of the topics over time and the
differences between the two regions' topics based on their similarity.
Finally, section~\ref{conclusion} summarizes our work and section~\ref{future} suggests some future work.

\section{Theoretic background}
\label{theoretic}
In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive texts body, machine learning 
researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents~\cite{Blei11introductionto}.

@@

When introducing the Latent Dirichlet Allocation in section~\ref{LDAexplained}, 
we will be using the terms prior and posterior probabilities
and likelihood function:

@@

where $\mathbb{E}$ is the expectation of the random variable $X$
~\footnote{Suppose random variable $X$ can take value $x_{1}$ with 
probability $p_{1}$, value $x_{2}$ with probability $p_{2}$, 
and so on, up to value $x_{k}$ with probability $p_{k}$. 
Then the expectation of this random variable $X$ is defined as
$\mathbb{E}[X] = x_{1}p_{1} + x_{2}p_{2} + ... + x_{k}p_{k}$.}.

@@

The probability distribution that the model uses is the multinomial, 
because it assumes that this is the distribution of topics in 
documents and words in topics (read Latent Dirichlet Allocation in section~\ref{LDAexplained}). 
In probability theory, the multinomial distribution is a generalization
of the binomial distribution. For example, it models the probability
of counts for rolling a $k$-sided dice $n$ times. For $n$ independent trials
each of which leads to a success for exactly one of $k$ categories, with
each category having a given fixed success probability, the
multinomial distribution gives the probability of any particular
combination of numbers of successes for the various
categories~\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words~\cite{Huang_maximumlikelihood}.

@@

The Dirichlet distribution is a way to model random probability mass functions
(PMFs)~\footnote{If you roll 1000 dice, the theoretical odds of any particular
number showing up (i.e. a 1, 2, 3, 4, 5, or 6) are 1/6. However, you won't get
that exact distribution in a real experiment due to manufacturing defects. If
you have ten dice, each die will have its own probability mass function (PMF).}
for finite sets. It is also sometimes used as a prior in Bayesian statistics.
The Dirichlet is the multivariate generalization of the beta distribution. It
is an extension of the beta distribution for modeling probabilities for two or
more events; when the result of the event has only 2 values, the Dirichlet
distribution is equal to the beta distribution.

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990~\cite{Deerwester90indexingby}. The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent
Semantic Indexing, or PLSI model~\cite{Hofmann:1999:PLS:312624.312649}. 
PLSI is a topic model where each word in a document is generated 
from a single topic which results in that each document in a corpus 
can be represented with a topic distribution. Later, in 2003 Blei et al. 
presented the Latent Dirichlet Allocation, or LDA~\cite{blei2003latent}. 
As opposed to PLSI, LDA is a statistically generative model for documents 
where each word in a document can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them~\cite{Deerwester90indexingby}. In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF~\footnote{TF-IDF stands 
for term frequency-inverse document frequency, and the TF-IDF weight is 
a weight often used in information retrieval and text mining. This weight 
is a statistical measure used to evaluate how important a word is to 
a document in a collection or corpus. The importance increases
proportionally to the number of times a word appears in the document (TF) but
is offset by the frequency of the word in the corpus (IDF).}. Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts~\cite{Salton:1988:TAA:54259.54260}.

@@

\begin{enumerate}
\item Pick a document \textbf{w} with probability $p($\textbf{w}$)$
\item For each of the $N$ words in \textbf{w}:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|$\textbf{w}$)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/PLSI_plate_notation.png}
\caption{Plate notation representing the PLSI model. \textbf{w} is the document 
index, $z$ is the topic drawn for the word $w$ from the topic distribution of 
the document}
\end{figure}
\end{center}
The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|$\textbf{w}$)$. Even though PLSI is generative for 
the analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data~\cite{blei2003latent}. Another problem is that the number 
of parameters in the model grows linearly with the number of documents.

\subsubsection{Latent Dirichlet Allocation (LDA)}
\label{LDAexplained}
LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics~\cite{Blei11introductionto}.

A limitation with PLSI is that there is no generative model for these
proportions, making it difficult to handle unseen documents. The reduced
document description result in PLSI is a vector where each element describes the
mixing proportion for a topic. The Latent Dirichlet Allocation model tries to
solve this limitation by setting a Dirichlet prior on the topic
distribution~\cite{blei2003latent}. In practice, this results in more reasonable
mixtures of topics in a document.

@@

The basic idea in LDA is that we define a topic to be a multinomial
distribution over a fixed vocabulary. Technically, the model assumes 
that the topics are generated first, before the
documents~\cite{Blei11introductionto}. Now for each document in the
collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit
multiple topics. Each document exhibits the topics with different
proportion (step 1); each word in each document is drawn from one of
the topics (step 2b), where the selected topic is chosen from the
per-document distribution over topics (step 2a). All the documents in
the corpus share the same set of topics, but each document exhibits
those topics with different proportion---that's the distinguishing
characteristic of LDA. A more detailed explanation of the model
follows in section~\ref{LDAmodel}.

@@

For each document $w$, in corpus $D$, the LDA generative process is as follows, 
with corresponding plate notation in Figure~\ref{figLDA}:

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section~\ref{oLDAvb}.

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously
~\cite{Myung:2003}.

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
~\cite{blei2003latent} alternative inference techniques use 
MCMC with Gibbs sampling~\cite{Griffiths02gibbssampling} and expectation 
propagation~\cite{Minka:2002:EGA:2073876.2073918}. The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed~\cite{Fox2011ATO}.

@@

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets
~\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

\paragraph{Online variational LDA}
\label{oLDAvb}
An online variational inference algorithm is proposed by Hoffman et all.~\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The \emph{online LDA} is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

We can see that unlike variational LDA, in online LDA we only need to
hold a small chunk of the data at a time, and once we're done
analyzing it, we never need it again. As with batch, once we've
estimated $\lambda$, we can find the most probable words for each
topic by looking at the word probabilities in each row of
$\lambda$~\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}.

@@

D. Blei's research focuses on probabilistic topic models, Bayesian methods and
approximate posterior inference. In his work \emph{Introduction to
Probabilistic Topic Models}~\cite{Blei11introductionto} 
he describes the basic ideas behind LDA and
applies the model to articles from the journal \emph{Science}. 
The model assumes that the topics are generated
first, before the documents and that the documents exhibit multiple
topics, but each one of them exhibit those topics in different
proportion. In our implementation we also use the LDA model and we
follow a similar approach, since we don't use any previous information
about the subjects that run through our dataset and the documents are
not labeled with keywords or topics. The topic distributions arise by
computing the hidden structure that resembles the thematic structure
of the collection.

The paper that mostly influenced our work is \emph{Finding Scientific Topics}, by
T. Griffiths and M. Steyvers~\cite{griffiths_steyvers04}. They investigated
abstract data from articles published in the Proceedings of the National Academy
of Sciences (PNAS) from 1991 to 2001 and compared research topics obtained from
topic modeling with existing categories. In our implementation, we also use 
the abstracts of the projects as documents for the model.

@@

Inspired by Griffiths and Steyvers, L. Sun and Y. Yin wrote a paper,
\emph{Discovering themes and trends in transportation research using topic
modeling}~\cite{Sun201749}, about topic modeling in transportation
research. Considering approximately 17000 abstracts of articles
published on 22 leading journals in transportation research as
documents, they infer 50 topics by using LDA. Compared to \emph{Finding
scientific topics}, Sun and Yin use more than one journal (22 popular
journals) and they investigate the journal-topic distribution and
journal similarity. In our analysis, we identify the EU and the USA as
2 regions (like 2 different journals), with different proportions of
documents-projects, and we visualize the topic similarity of the two
regions per framework programme.

@@

\section{Methodology}
\label{methodology}
In this section, we explain the procedure that we followed in order to
apply the LDA model in our dataset. It consists of information about 
EU projects funded for research and innovation under the FP4--H2020 
and the awards/projects that got funded by NSF from 1994--2017 in USA. We
consider the projects' description/abstract as the documents in LDA.

@@

Below, we provide the general topics of the EU projects as listed in the EU 
Open Data Portal~\footnote{https://data.europa.eu/euodp/en/data}:
\begin{itemize}
\item Agriculture, Fisheries and Food
	\begin{description}
	\item Agriculture, Food safety, Maritime affairs and fisheries
	\end{description}
\item Business
	\begin{description}
	\item Competition, Enterprise, Single market, Trade
	\end{description}
\item Culture and Education
	\begin{description}
	\item Audiovisual and media, Culture, Education training and youth, Multilingualism
	\end{description}
\item Customs and Tax
	\begin{description}
	\item Customs, Taxation
	\end{description}
\item Development and Humanitarian aid
	\begin{description}
	\item Development and cooperation, Humanitarian aid and Civil protection, Human rights
	\end{description}
\item Economy and Finance
	\begin{description}
	\item Budget, Economic and monetary affairs, Fraud prevention
	\end{description}
\item Employment and Social affairs
\item Enlargement and Foreign affairs
\item Environment and Energy
	\begin{description}
	\item Climate action, Energy, Environment
	\end{description}
\item EU institutions
	\begin{description}
	\item Institutional affairs
	\end{description}
\item Health
	\begin{description}
	\item Health, Sport
	\end{description}
\item Justice and Citizens' rights
	\begin{description}
	\item EU citizenship, Consumers, Justice and Home affairs
	\end{description}
\item Regions and Local development
	\begin{description}
	\item Regional policy
	\end{description}
\item Science and Technology
	\begin{description}
	\item Space, Digital Economy and Society, Research and Innovation
	\end{description}
\item Transport and Travel
\end{itemize}

The data that are provided by CORDIS (via the EU Open Data Portal) are
open, free and can be used in research, applications, commercial or
non-commercial purposes. The dataset that we chose for the following
analysis contains information about all the projects funded by the
European Union from 1994 to 2017. These projects got approved under
the framework programme (FP4--H2020) for research and technological
development. Grant information is provided for each project, 
including reference, acronym, dates, funding, programmes, 
participant countries, subjects and objectives.

@@

\begin{itemize}
\item FP4: fourth framework programme (1994--1998)
\item FP5: fifth framework programme (1998--2002)
\item FP6: sixth framework programme (2002--2006)
\item FP7: seventh framework programme (2007--2013)
\item H2020: Horizon 2020 framework programme (2014--2020)
\end{itemize}

@@

\emph{Management software \& systems} is a subject that appears only in FP4 and
FP5, with the 12\% of the projects. \emph{Information systems \& technologies'}
related projects cover the 8,5\% of the projects on average from FP5--H2020.
Since technology has an impact in every aspect nowadays, it is normal to have a
stable popularity over the years.
